### 2020-07-12 response/response-demography-3.tex

\begin{center} \textbf{\large Birth Spacing in the Presence of Son
Preference and 
Sex-Selective Abortions: India's Experience over Four Decades}
\end{center}

\begin{center} Response to Editor and Reviewer Comments \end{center}

\noindent Please find attached the revised version of my paper,
``Birth Spacing in the Presence of Son Preference and Sex-Selective
Abortions:
India's Experience over Four Decades.''

[Other changes]
\begin{enumerate}
	\item Rewritten Estimation Strategy section
	\begin{enumerate}
		\item Clarify proportionality assumption
		\item Better explain set-up of model and transformation of data needed
		\item Remove discussion of explanatory variables so they are
		only covered once (in the ``Data'' section)
		\item Better explanation of how I calculate measures used for the spacing
		analysis
	\end{enumerate}
	\item Changes in Data section
	\begin{enumerate}
		\item Lead with what birth intervals I cover right after introducing
		data
		\item Move restriction to only Hindus to be the first restriction
		I discuss since that is the most important
	\end{enumerate}
\end{enumerate}

@@

This grouping also allows me to capture the U-shape in female labor force participation.
The upper level of education could be split into having graduated secondary and having 
graduated with a university degree since women with a completed university degree have 
a higher labor force participation, but the sample of university graduates is too small 
to allow for estimations, especially for the first three rounds of the surveys.
Furthermore, with four groups there is already a substantially number of results to 
present.

\item The word ``spell'' is introduced with no definition. Is spell the
same as birth interval? Is it measured in months? This information would
be useful for the reader.

[Response:]

I now define spell as the unit of analysis---the period from one parity birth to 
the following birth or censoring---when I first use the word in the ``Estimation 
Strategy'' section.
Furthermore, I clarify that for estimation purposes spells begin nine months 
after the previous birth since this is the earliest we should expect to observe 
a new birth.
This duration is equivalent to the interpregnancy interval, which is often used in 
the literature on spacing and child mortality.

To make comparison with both the NFHS reports and the prior literature on
birth spacing easier and avoid confusion, I have rewritten the discussions of 
spacing and redone the graphs and table, so all results now reflect birth 
intervals in months.
I still retain the use of ``spell'' to indicate a generic period from one parity 
birth to another, i.e. second spell is from first birth to the second birth or
censoring, in the graphs/tables.

In the Data section, I have also changed the discussion of the imposed censoring 
clearer by specifying the cut-offs for both the spell duration and the birth interval.
In the ``How Birth Spacing Changed'' section, I explain that I add nine month
to the estimated spell length to get to birth intervals.
In the ``Mortality and the Changing Birth Spacing'' section, I retained the
definition of the periods but updated the numbers and the discussion of
results so they reflect birth intervals rather than pregnancy intervals.

### 2020-07-13 paper/sex_selection_spacing-ver4.tex
Add background/questions to Conclusion; begin data quality discussion

In addition to a large number of women surveyed and the long period covered, a
benefit of the NFHS is that enumerators pay careful attention to the spacing 
between births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, remains a potential problem.
Recall error is heavily dependent on how long ago a woman was married, and I, therefore, 
drop women married 22 years or more based on the discussion in the Appendix.

@@

% What has happened to spacing and why

One effect of the significant changes that India has experienced over the last four 
decades is that spacing between births has increased.
The size of the increase is increasing in parity and the mother's
education and is larger in urban than in rural areas.

The most substantial increases in birth intervals arise, however, with
heavy use of sex selection. Some of these increases are so large that we
even observe a reversal of the traditional spacing pattern where strong
son preference was associated with shorter birth intervals, the fewer
sons a family has \citep{ben-porath76b,Leung1988}.

Despite the longer average birth intervals, a large number of women
still have very short birth intervals. 
Hence, the changes in birth
spacing are driven by the longer intervals becoming even longer over
time, both for women who do not use sex selection, but especially for
those who do and end up with multiple abortions. 
The exception to the
small change in short spacing is the delay in conception for women with
very high use of sex selection, presumably because this group knows that
the next child will be a boy.

% fertility
Fertility has declined for all groups, but the likelihood of having an additional child
still depends strongly on the number of sons, with women with no sons having the highest 
probabilities.

That those women most likely to use sex selection are also the most
likely to have a next birth, combined with the secular increases in
spacing, makes the total fertility rate a more biased measure of cohort
fertility. 
This bias was most prominent early in the spread of sex
selection when the fertility rate was up to one child lower than the
predicted cohort fertility. 
It is, however, still present with the
predicted cohort fertility ten to twenty percent higher than the
fertility rate.

Tempo effects are studied extensively in the literature \citep[See, for example, ][]{Bongaarts1999}.
Still, there are, to my knowledge, no other cases where there has been as 
substantial an increase in birth intervals and associated bias in fertility rates 
as for India.
It is conceivable that we might see increases in TFR as birth spacing 
stabilizes or even shortens again if interventions against sex selection are successful.

@@

An implication of the results here is that population growth in India
will be higher than expected. There has been a rapid reduction in infant
mortality risk, the longer spacing should further reduce mortality risk,
and cohort fertility is higher than previously realized.

@@

There are two critical questions that future research should address.
Sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection could be an
unintended side-effect, rather than a result of girls becoming more
valued as is often assumed. 
Comparing prior children's outcomes across
sex composition and the sex of the next child could be a way to
understand why girls' health outcomes improve in the presence of sex selection.

>>>>

In addition to a large number of women surveyed and the long period covered, a significant 
benefit of the NFHS over other surveys is that enumerators pay careful attention to the 
spacing between births and probe for ``missed'' births.%
\footnote{
The method used to probe for missing births depends on the survey.
NFHS-1 probe for each calendar birth interval that is four or more years.
NFHS-2 asked for stillbirths, spontaneous and induced abortions, and also probed 
for each calendar birth interval four or more years.
NFHS-3 and NFHS-4 did not directly use birth intervals, but asked whether there were any 
other live births between (name of previous birth) and (name), including any children who 
died after birth, and asked for births before the birth listed as first birth and
after the last birth listed as the last birth.
}
The quality of the birth date data is especially important here since the reliability of the 
results depends on the correctness of the birth histories provided by the respondents.
The main concern is underreporting of deceased children, especially a systematic recall error 
where respondents' likelihood of reporting the birth of a deceased child depends 
on the sex of that child.
To establish the quality of the birth interval data, I provide a detailed discussion of the 
degree of systematic recall error in the Online Appendix.
Recall error appear heavily dependent on how long ago a woman was married, 
and I, consequently, drop women married 22 years or more.

@@

% Background
Over the last four decades India saw a dramatic increase in the male-to-female
sex ratio at birth as access to sex selection spread and son preference remained. 
At the same time, economic growth was strong, schooling increased, and the total
fertility rate fell to close to replacement.

% Questions
The main question I address in this paper is how birth spacing responded to the major changes
that occurred in India over the last four decade, in particular the spread of sex selection.
In addition, I examine two related questions.
First, did the changes in birth spacing bias the standard fertility estimates for India? 
Second, what is the relationship between infant mortality and the changes in birth spacing 
and sex selection?

% What has happened to spacing and why
Birth intervals has lengthen, often substantially, over the last four decades.
The increases in spacing are larger, the higher the parity and the higher the mother's
education, and is larger in urban than in rural areas.

Groups with the highest use of sex selection have the most substantial increases in 
birth intervals, with increases in average birth intervals of up to 12 months. 
Some of these increases are so large that we even observe a reversal of the traditional 
spacing pattern: instead of the absence of sons leading to the shortest birth intervals, 
son preference and sex selection lead some groups to have longest birth intervals when
they have no sons.

Despite the longer average birth intervals, a large number of women still have very 
short birth intervals. 
Hence, the changes in birth spacing are driven predominately by the longer intervals 
becoming even longer over time.
This pattern holds generally, but is especially prevalent for those who use sex selection 
and, therefore, may end up with multiple abortions. 
The exception to the small change in short spacing is the delay in conception for women 
with very high use of sex selection, presumably because this group knows that
the next child will be a boy.

% fertility
Fertility has declined for all groups, but the likelihood of having an additional child
still depends strongly on the number of sons, with women with no sons having the highest 
parity progression probabilities.

The combination of the secular increases in spacing and the fact those women most likely to use
sex selection are also those most likely to have a next birth, makes the total fertility rate a 
more biased measure of cohort fertility. 
This bias was most prominent early in the spread of sex selection, when the fertility rate 
was up to one child lower than the predicted cohort fertility. 
It is, however, still present, with the predicted cohort fertility ten to twenty percent 
higher than the fertility rate.

Tempo effects are studied extensively in the literature \citep[See, for example, ][]{Bongaarts1999}.
Still, there are, to my knowledge, no other cases where there has been as 
substantial an increase in birth intervals and associated bias in fertility rates 
as for India.
It is conceivable that we might see increases in the total fertility rate as birth spacing 
stabilizes or, if interventions against sex selection are successful, even shortens again.

@@

An important implication of the results here is that population growth in India
may be higher than expected. 
There has been a rapid reduction in infant mortality risk, the longer spacing should 
further reduce mortality risk, and cohort fertility is higher than previously realized.

@@

There are two critical questions that future research should address.
First, sex selection means that girls-only families are less likely to have
very short birth intervals, which may reduce sibling competition. 
Hence, better health outcomes for girls with sex selection could be an
unintended side-effect, rather than a result of girls becoming more
valued as is often assumed. 
Comparing prior children's outcomes across sex composition and the sex of the next child 
could be a way to understand why girls' health outcomes improve in the presence of sex selection.
